---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Benjamin Egan"
date: "2-17-25"
output: github_document
---

```{r load-packages, warning = FALSE, message = FALSE, echo = FALSE}
library(tidyverse) 
library(dsbox)
library(ggplot2)
library(mosaicData) 
staff <- read_csv("data/instructional-staff.csv")
```


link to assignment: https://datascience4psych.github.io/DataScience4Psych/lab06.html

### Exercise 1

```{r wide to long}

staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))

```

### Staff by Year Graph

```{r graph, fig.width = 8}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line()+
  theme_bw()+
  labs(
    x = "Year",
    y = "Percent of Total Instructional Staff",
    title = "Trends in Instructional Staff Employee Status",
    color = "Faculty Status"
  )
```


### Fisheries Graph

```{r Fisheries}
fisheries <- read_csv("data/fisheries.csv")

view(fisheries)
```

Looking at this data set, we have four variables. Country, capture, aquaculture, and total. Based on their data, it's clear that China leads aquaculture by a landslide. One way we could visualize this is by taking out China and individually reporting their statistics. It would help with the visual. Another way to help would be to create two separate graphs, one for aquaculture and one for capture. The third way (the one I am picking to visualize) is to show the top 10 fisheries in the total column. From there, I would indicate the proportion of the total fish obtained via aquaculture and capture. There is probably a better way to visualize it, but for now I overlay a bar plot of the amount obtained by aquaculture with a blank bar plot of the total amount of fish (to give the illusion of the rest being by capture.)

```{r visual for fisheries, fig.width = 10}

top_10 <- fisheries %>%
  arrange(desc(total)) %>%
    slice(1:10) 


top_10 %>% ggplot(aes(
  x = fct_reorder(country, total),
  y = total
)) +
  geom_bar(stat = "identity", aes(y = aquaculture), alpha = .6, fill = "salmon")+
  geom_bar(stat = "identity", fill = NA, color = "Black")+
  coord_flip()+
  labs(
    x = "Country",
    y = "Amount in Tons",
    title = "Tonnage of Fish captured",
    subtitle = "Salmon indicates the proportion farmed via aquaculture instead of capture"
  ) 

```


## Mosaic Data

```{r load libraries, echo = FALSE}
data("Whickham")
```

### Q/A answers

```{r Q/A answers, message = FALSE, fig.width = 12}
#Smoking Visual
Whickham %>%
ggplot(aes(
   x=age,
   fill = fct_rev(smoker)
   )) + 
    geom_histogram(color = "black")+
  theme_bw()+
  labs(
    x = "Age",
    y = "",
    title = "Smokers and non smokers at baseline",
    fill = "Are they a smoker?"
  )+
    theme(plot.title = element_text(size = 20),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 15),
        )

#Death Visual
Whickham %>%
ggplot(aes(
   x=age,
   fill = fct_rev(outcome)
   )) + 
    geom_histogram(color = "black")+
  theme_bw()+
    labs(
    x = "Age",
    y = "",
    title = "Survival status after 20 years",
    fill = "Are they still alive?"
  )+
    theme(plot.title = element_text(size = 20),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 15),
        )

#create distinguishing variable
Whickham_alter <- Whickham %>%
  mutate(Specific = case_when(
    smoker == "Yes" & outcome == "Dead" ~ "Dead smoker",
    smoker == "Yes" & outcome == "Alive" ~ "Alive smoker",
    smoker == "No" & outcome == "Dead" ~ "Dead non-smoker",
    smoker == "No" & outcome == "Alive" ~ "Alive non-smoker"
      ))

#visual for all groups
Whickham_alter %>%
ggplot(aes(
   x=age,
   fill = Specific
   )) + 
    geom_histogram(color = "black")+
  theme_bw()+
  facet_wrap(~Specific)+
    labs(
    x = "Age",
    y = "",
    title = "Survival status after 20 years",
    fill = "Survival status and baseline smoking"
  )+
    theme(plot.title = element_text(size = 20),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 15),
        )

```

1. This was an observational study, since they weren't trying out an intervention for smoking.

2. There were 1314 observations in this data set (all women)

3. There are three variables; outcome, smoker, and age. Outcome is a categorical variable with answers (Alive or Dead). Smoker is a categorical variable with answers (No or Yes). Age is a continuous numerical variable that recorded their age in years.

4. Based on the two simple graphs I made, I would expect that smoking decreasing your health, leading to earlier death. The more complex graph shows me that there is a more spread out distriubtion of death in the smokers than htere is for the non-smokers.

### A different visual

```{r visuals asked for}
Whickham %>%
  count(smoker, outcome) %>%
  group_by(outcome) %>%
  mutate(percent_outcome = (n / sum(n))*100)

ggplot(Whickham, aes(y = smoker, 
                     fill = outcome)) +
  geom_bar(position = "fill") + 
  labs(title = "Smoking outcome",
       y = "Did they smoke?", x = NULL)

```

```{r age breakdown, fig.width= 10}
Whickham_age_breakdown <- Whickham_alter %>%
 mutate(age_cat = case_when(
    age <= "44" ~ "18-44",
    age > "44" & age <= "64" ~ "45-64",
    age > "64" ~ "65+",
      ))

Whickham_age_breakdown %>%
 ggplot(aes(
   y = fct_rev(smoker), 
   fill = fct_rev(outcome))) +
  geom_bar(position = "fill") + 
   facet_wrap(~age_cat)+
  labs(title = "Smoking outcome",
     y = "Did they Smoke?", x = NULL)

```

Have to say, it's hard to see the difference in this graph. I included the table below. It looks like there are more people alive longer based on the counts for the 65+ category.

```{r table extension}

Whickham_age_breakdown %>%
  count(age_cat, smoker, outcome) %>%
  group_by(outcome) %>%
  mutate(percent_outcome = (n / sum(n))*100)

```

## Additional Plot

A suggestion for how to make the visual easier to see? You can more clearly see in this plot that the distribution of people who's status is different based on smoking status. Smokers have a more even distribution while non-smokers are clearly skewed.

```{r additional plot, warning = FALSE, fig.width = 10}
Whickham_age_breakdown %>%
ggplot(aes(
   x=age_cat,
   fill = Specific
   )) + 
    geom_histogram(stat = "count", color = "black")+
  theme_bw()+
  facet_wrap(~Specific)+
    labs(
    x = "Age",
    y = "",
    title = "Survival status after 20 years",
    fill = "Survival status and baseline smoking"
  )+
    theme(plot.title = element_text(size = 20),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 15),
        )
```
